Efficient pedigree recording for fast population genetics simulation

### 2017-11-16 forwards_paper.tex
did through discussion

\paragraph{Sequential simplification}
At any point the simulation scheme above, genealogical history is recorded
in a tree sequence. This has two additional advantages. First, simplification
can be run periodically through the simulation, taking the set of samples to be
the entire currently alive population. This is important as it keeps memory
usage from growing linearly (and quickly) with time. Second, the simulation can
be \emph{begun} with a tree sequence produced by some other method -- for
instance, by a coalescent simulation with \msprime. This allows for
incorporation of deep-time history beyond the reach of individual-based
simulations. Since geographic structure from times longer ago than the mixing
time of migration across the range has limited effect on modern genealogies
\citep{wilkins_separation} (other than possibly changing effective population
size \citet{durretspatial}), this may not negatively affect realism.

@@

\plr{We could compute the recursions for these quantities explicitly for this simple WF case...}

@@

%%%%%%%%%%%%%%%%%%%%%%
\section*{Conclusion}

This works great and very fast.

Having results in a tree sequence is really good downstream too:
The tree sequence encoding is compact but it is also efficient to process.
Many algorithms to compute statistics of interest for population genetics
are naturally expressed in terms of tree topologies. For example, the
pairwise nucleotide diversity $\pi$, is defined as the average number of
differences between sequences in the sample. Computing $\pi$ directly
from observed sequence data requires $O(n^2 m)$ time, since we must
compare all pairs of samples at every site. However, given the topology of
the trees at every site and the locations of mutations much better algorithms
are possible. Using the fact that we can count the number of samples below
a given node efficiently using previously described tree sequence
algorithms~\citep{kelleher2016efficient}, the time required to compute $\pi$
becomes roughly $O(n \log n + m)$.
The \msprime\ API provides a method to compute $\pi$ among arbitrary subsets of the
samples in a tree sequence, and calculating $\pi$ over all 500K samples
in the example above required about 1.2 seconds.
\jk{I haven't checked the analysis of computing $\pi$ here and not thought about
it too deeply. I guess we should also compare this with something else in order to
show that this time is pretty good. We could get a subset of the sites and push
the data into pylibseq maybe, or scikit-allel??}

This is a general-purpose strategy that can be applied to other methods.

A hybrid approach, in which the recent history of a large sample of
individuals is simulated under a detailed forwards-in-time model and the deep history
under the coalescent, is an attractive prospect.
\jk{Not sure where this should go; perhaps in the discussion? We can make the point that
by keeping track of the genealogies forwards in time we can easily combine these with the
deep history simulated by msprime}

Parallelization will be important;
this requires only one-way communication.

>>>>

\paragraph{Sequential simplification}
At any point the simulation scheme above, genealogical history is recorded
in a tree sequence. This has two additional advantages. First, simplification
can be run periodically through the simulation, taking the set of samples to be
the entire currently alive population. This is important as it keeps memory
usage from growing linearly (and quickly) with time. Second, the simulation can
be \emph{begun} with a tree sequence produced by some other method -- for
instance, by a coalescent simulation with \msprime.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}

In this paper, we have shown that storing pedigrees
and associated recombination events
in a forwards-time simulation
not only results in having available a great deal more information about the simulation,
but also can speed up the simulation by orders of magnitude.
To make this feasible, 
we have described how to efficiently store this information in tables,
and have described a fundamental algorithm for simplificication of tree sequences.
Conceptually, recording of genealogical and recombination events
can happen independently of the details of simulation;
for this reason, we provide a well-defined and well-tested API in C and in python
for use in other code bases.

The tree sequences produced by default by this method
are very compact, storing genotype \emph{and} genealogical information
in a small fraction of the space taken by a compressed VCF file.
The format is also highly efficient to process,
leading to advantages for downstream analysis as well.
This is because many algorithms to compute statistics of interest for population genetics
are naturally expressed in terms of tree topologies,
and so can be quickly computed from the trees underlying the tree sequence format. 
For example, pairwise nucleotide diversity $\pi$, is the average density of
differences between sequences in the sample. 
To compute this directly from sequence data at $m$ sites in $n$ samples
requires computing allele frequencies, taking $O(nm)$ operations.
By using the locations of the mutations on the marginal trees,
and the fact that these are correlated, 
dynamic programming algorithms similar to those in~\citep{kelleher2016efficient}
can do this in roughly $O(n + m \log n)$ operations.
The \msprime\ API provides a method to compute $\pi$ among arbitrary subsets of the
samples in a tree sequence, which took about 1.2 seconds
applied to the example tree sequence above with 1.1 million mutations 
in 200 megabases for 500,000 samples.

Another attractive feature of this set of tools
is that it makes it easy to incorporate \emph{prior history},
simply by seeding the simulation with a (relatively inexpensive) coalescent simulation.
This allows for incorporation of deep-time history beyond the reach of individual-based simulations. 
Since geographic structure from times longer ago than the mixing
time of migration across the range has limited effect on modern genealogies
\citep{wilkins2004separation} (other than possibly changing effective population
size \citet{barton2002neutral,cox2002stepping}), this may not negatively affect realism.
